---
title       : Sets and relations
subtitle    :
author      : Paul Schrimpf
job         :
date        : "`r format(Sys.time(), '%d %B, %Y')`"
output:
    revealjs::revealjs_presentation:
        self_contained: true
        theme: league
        transition: slide
        center : true
        highlight : zenburn
        reveal_options:
            slideNumber: false
            center : true
            help : true
            previewLinks: true

    ioslides_presentation :
        self_contained: true
        code_folding: hide
        lib_dir : deps
        theme : journal
        fig_width: 8
        fig_height: 6

#ext_widgets: {rCharts: [libraries/nvd3]}
## For interactive chart to work, must either publish or view from
## local webserver, else CDN links won't work
---

# Sets

## Definition & Notation

- Well-specified collection of elements
- set $A$, element $a \in A$
- Example:
$$
  A = \{4, 5, 6 \} = \{ n \in \mathbb{N} : 3 < n < 7 \}
$$

## Operations & relations

- **Union:** $A \cup B = \{x: x\in A \text{ or } x \in B\}$.
- **Intersect:** $A \cap B = \{x: x \in A \text{ and } x \in
  B\}$.
- **Minus:** $A \setminus B = \{ x: x\in A \text{ and }
  \not\in B \}$
- **Product:** $A \times B = \{(x,y): x \in A, y \in B \}$
- **Power set:** $\mathcal{P}(A) =$ set of all subsets of $A$
- **Subset** $A \subseteq B$ means that if $a\in A$ then $a \in B$
  + $A \subset B$ means $A \subseteq B$ and $\exists b \in B$ such
  that $b \not\in A$


# Relations



## Relations

- **relation** on two sets $A$ and $B$ is any subset of $A
  \times B$, $R \subseteq A \times B$. We usually denote relations by
  $a \sim_R b$ if $(a,b) \in R$
  + Example: $A=B=\mathbb{R}$, $<$ is associated with $R_{<} = \{(a,b)
  \in \mathbb{R}^2 : a<b \}$



## Equivalence Relation

- **equivalence relation** is
  + relexive: $a \sim_R a$ $\forall a \in A$
  + transitive: $a \sim_R b$ and $b \sim_R c$ implies
    $a \sim_R c$
  + symmetric: $a \sim_R b$ implies $b \sim_R a$,
- partitions $A$ into equivalence classes, $\sim(x) = \{a \in A: a
  \sim x \}$



## Weak order

- $\preceq$ is a **weak order** if
  + relexive: $a \preceq a$ $\forall a \in A$
  + transitive: $a \preceq b$ and $b \preceq c$ implies
    $a \preceq $c
  + complete: for any $a, b \in A$, either $a \preceq b$ or
  $b \preceq a$ (or both)
- Example: preference relation



## Pareto order

- $n$ individuals, preferences $\succeq_i$
- Pareto order:
$$
x \succeq^P y \text{ if } x \succeq_i y \text{ for all } i=1,...,n
$$
- transitive? reflexive? complete?


##

```{r, echo=F, fig.width=6, fig.height=2.2}
library(ggplot2)
library(gridExtra)
library(scales)
outcomes <- 5
n <- 3
ranking <- list()
Rpref <- list()
ranking[[1]]  <- c(1,2,3,4,5)
ranking[[2]]  <- c(2,3,1,4,5)
ranking[[3]]  <- c(4,3,1,2,5)
for (i in 1:n) {
  Rpref[[i]] = matrix(F, nrow=outcomes, ncol=outcomes)
  for (j in 1:(outcomes-1)) {
    r = ranking[[i]][j]
    Rpref[[i]][r,ranking[[i]][(j+1):outcomes]] = T
  } 
}

plotpref <- function(R) {
  outcomes  <- nrow(R)
  df  <- expand.grid(1:outcomes, 1:outcomes)
  df$R  <- as.vector(R)
  fig  <-  ggplot(data=df, aes(x=Var1, y=Var2, fill=R)) +
    geom_tile(width=0.9, height=0.9, ) +
    theme_bw()+ 
    theme(axis.title.x=element_blank(),
          axis.title.y=element_blank(),
          legend.position="none") 
  fig
}
fig <- list()
for (i in 1:n) {
  fig[[i]]  <- plotpref(Rpref[[i]]) + labs(title=sprintf("Person %d",i),
                                           subtitle=paste("Preference order:", paste(ranking[[i]], collapse=",")))
}
grid.arrange(grobs=fig, ncol=3)
```

```{r, echo=F,  fig.width=2, fig.height=2.2}
Rpareto = Rpref[[1]]
for (i in 2:n) {
  Rpareto[Rpareto != Rpref[[i]]]  <-  NA
}
figp  <- plotpref(Rpareto) + labs(title="Pareto Order")
figp
```

## Social choice rule

- Combines preferences into single weak order $F(\succeq_1, ...,
\succeq_n) = \succeq$
- Completes Pareto order if $x \succeq^P y$ implies $x \succeq y$

## Examples

- Binary voting
  + social choice rule is determined by majority vote on each pair
   $$ x \succeq y \iff x \succeq_i y \text{ for at least $(n+1)/2$
  individuals } $$

---

```{r, echo=F, fig.width=6, fig.height=2.2}
grid.arrange(grobs=fig, ncol=3)
```

```{r, echo=F, fig.width=4, fig.height=2.2}
Rbin  <- matrix(0, nrow=outcomes, ncol=outcomes)
for(i in 1:n) {
  Rbin  <- Rbin + Rpref[[i]]
}
Rbin  <-  Rbin>=(n+1)/2
figb <- plotpref(Rbin) + labs(title="Binary voting")
grid.arrange(grobs=list(figp,figb), ncol=2)
```


## Examples

- Majority voting
  + Each person votes for most preferred outcome
  + Social choice rule ranks by number of votes
  + Ties broken by number of votes for second most preferred, then third, etc


---
```{r, echo=F, fig.width=6, fig.height=2.2}
grid.arrange(grobs=fig, ncol=3)
```

```{r, echo=F, fig.width=6, fig.height=2.2}
votes <- matrix(0,nrow=outcomes,ncol=outcomes)
for (i in 1:n) {
  for (c in 1:outcomes) {
    f <- ranking[[i]][c]
    votes[f,c] <- votes[f,c] + 1
  }
}
Rmv  <- matrix(NA, nrow=outcomes, ncol=outcomes)
for(i in 1:outcomes) {
  for (j in 1:outcomes) {
    for (c in 1:outcomes) {
      if (votes[i,c] != votes[j,c]) {
        Rmv[i,j] = votes[i,c]>votes[j,c]
        break
      }
    }
  }
  Rmv[i,i] = F
}
figv <- plotpref(Rmv) + labs(title="Majority voting")
grid.arrange(grobs=list(figp,figb, figv), ncol=3)
```


## IIA

- **Independence of irrelevant alternatives** if for every $A \subseteq X$, if $\{\succeq_i\}_{i=1}^n$ and
$\{\succeq_i'\}_{i=1}^n$ are two sets of individuals preferences that agree on $A$,
\[ x \succeq_i y \iff x \succeq_i' y \; \forall x,y \in A \text{ for each
} i\]
then $\succeq=F(\succeq_1, ..., \succeq_n)$ and $\succeq'=F(\succeq_1', ...,
\succeq_n')$ also agree on $A$


## Arrow's impossibility theorem

It is impossible to complete the Pareto order in a way that
is not dictatorial and is independent of irrelevant alternatives when
$|X| \geq 4$


# Proof of Arrow's impossibility theorem

## Decisive

- With social choice rule $F$, a group of individuals, $S \subseteq
\{1,\cdots,n\}$ is **decisive** over $x, y \in X$ if for any
individual preference relations if $x \succ_i y$ for all $i \in S$
implies $x \succ y$, where $\succ=F(\succ_1,..., \succ_n)$.

## Field expansion lemma

- Assume $\exists$ $x$ and $y$ such that some group $S$ is decisive
over $x$, $y$
- Use IIA to show $S$ decisive over all other $z$ and $w$

## Group contraction lemma

- If $S$ decisive and $|S| > 1$, then $\exists S' \subset S$ such
  that $S'$ decisive


# Cardinality



## Bijection

$$ f: A \to B $$
is

- **one-to-one** (injective) if for $b \in B$ there is at most one
  $a \in A$ such that $f(a) = b$
- **onto** (surjective) if $\forall b \in B=$ $\exists a \in A$
  s.t. $f(a) = b$
- **bijective** if both



## Cardinality

- **Cardinality** of $A$, $|A|$
  + $|A| = |B|$ if $\exists$ bijection between $A$ and $B$
  + **finite** if $|A| = |\{1, 2, ..., n\}| = n$
  + **countable** if $|A| = |\mathbb{N}|$
  + **uncountable** otherwise




## Integers

- $\mathbb{Z}$ is ???



## Rationals

- $\mathbb{Q}$ is ???



## Reals

- $\mathbb{R}$ is ???

