---
title       : "Optimal Control"
subtitle    : 
author      : Paul Schrimpf
job         : 
date        : September 2017
output:
  revealjs::revealjs_presentation:
    self_contained: false
    theme: league
    transition: slide
    center : true
    highlight : zenburn
    reveal_plugins: ["chalkboard"]
    reveal_options:
      slideNumber: false
      center : true
      help : true
      previewLinks: true
      chalkboard:
        theme: whiteboard
      #  toggleNotesButton: true
      #  toggleChalkboardButton: true
#ext_widgets: {rCharts: [libraries/nvd3]}
## For interactive chart to work, must either publish or view from
## local webserver, else CDN links won't work 

## To create an html file from this, in R enter "library(revealjs); render("optimalControl.Rmd")
--- 

# Motivating examples

## Optimal growth

$$
\begin{aligned}
  \max_{c(t),k(t)} & \int_0^\infty e^{-\delta t} u(c(t)) dt \\
  \text{ s.t. } & \frac{dk}{dt} = f(k(t)) - \phi k(t) - c(t) \\
  & k(0) = k_0 \\
  & 0 \leq c(t) \leq f(k(t))
  \end{aligned}
$$

---

## Investment with adjustment costs

$$
  \begin{aligned}
    \max_{k(t),i(t),l(t)} & \int_0^T e^{-r t} \left[p f(k(t),l(t)) - qi(t) -
    wl(t) -
    c(i(t),k(t)) \right] dt \\
  \text{ s.t. } 
    & \frac{dk}{dt} = i(t) - \delta k(t) \\
    & k(0) = k_0 \\
    & k(t) \geq 0 \\
    & l(t) \geq 0
  \end{aligned}
$$

---

## Contracting with a continuum of types

$$ 
\begin{aligned}
\max_{q(\theta),T(\theta)} & \int_{\theta_l}^{\theta_h} 
\left[T(\theta) - cq(\theta)\right]
f_\theta(\theta) d\theta \notag \\
& \text{s.t.} \notag \\
&\theta \nu\left(q(\theta)\right) - T(\theta) \geq 0  \forall
\theta \label{pc} \\
&\theta \nu\left(q(\theta)\right) - T(\theta) \geq
\max_{\tilde{\theta}} \theta \nu\left(q(\tilde{\theta}) \right) -
T(\tilde{\theta}) \forall \theta \label{ic} 
\end{aligned}
$$

---

## Optimal income taxation
$$
  \begin{aligned}
    \max_{\ell,t} &  \int_{w_l}^{w_h} G\left(u(w\ell(w) - t(w\ell(w)),\ell(w))
    \right) f(w) dw \\
    & \text{ s.t. } \\
    & \int_{w_l}^{w_h} t(w) f(w) dw \geq g \\
    & \ell(w) \in \argmax_{\tilde{\ell}} u(w\ell - t(w\ell), \ell )
  \end{aligned}
$$

<!--- ##################################################### --->

# The maximum principle

## Generic optimal control problem

$$ 
\begin{aligned}
  \max_{x(t),y(t)} & \int_0^T F(x(t),y(t),t) dt \\
  & \text{ s.t.} \\
  & \frac{d y}{dt} = g(x(t),y(t),t) \forall t \in [0,T] \\ 
  & y(0) = y_0
\end{aligned}
$$

--- 

## Discretization

> - Discretized problem
$$
\begin{aligned}
  \max_{x_1, ..., x_J, y_1, ..., y_J} & \sum_{j=1}^J \Delta F(x_j,
  y_j,\Delta j) 
  & \text{ s.t.} \\
  & y_j - y_{j-1} = \Delta g(x_j,y_j,\Delta j) \text{ for} j = 1,...,J
\end{aligned}
$$

--- 

## Discretization

> - Discrete first order conditions
$$
\begin{aligned}
  [x_j]: &&  \frac{\partial F}{\partial x} + \lambda_j \frac{\partial
    g}{\partial x} = & 0 \\ 
  [y_j]: &&  \frac{\partial F}{\partial y} + \lambda_j 
  \frac{\partial g}{\partial y} = & -\frac{\lambda_{j+1} -
    \lambda_j}{\Delta} \\
  [\lambda_j]: && g(x_j,y_j,\Delta j) = & \frac{y_j - y_{j-1}}{\Delta}.
\end{aligned}
$$

--- 

## Discretization

> - Limit as $\Delta \to 0$
$$ 
\begin{aligned}
  [x_j]: &&  \frac{\partial F}{\partial x} + \lambda(t) \frac{\partial
    g}{\partial x} = & 0 \\ 
  [y_j]: &&  \frac{\partial F}{\partial y} + \lambda(t)
  \frac{\partial g}{\partial y} = & -\frac{d\lambda}{dt} \\ 
  [\lambda_j]: && g(x_j,y_j,\Delta j) = & \frac{dy}{dt}
\end{aligned}
$$

---

## Functional directional derivatives

> - $Q(x,y) \equiv  \int_0^T F(x(t),y(t),t)dt$
> - Directional derivative, for $v:[0,T] \to \mathbb{R}$ and $w:[0,T]
    \to \mathbb{R}$, 
    $$ dQ(x,y;v,w) = \lim_{\alpha \to 0} \frac{Q(x+\alpha v, y +
\alpha w) - Q(x,y)}{\alpha} 
    $$

--- 

## Lagrangian

$$
\begin{aligned}
L(x,y,\lambda,\mu_0) = & \int_0^T F(x(t),y(t),t) dt - \\ & - \int_0^T
  \lambda(t)\left( \frac{dy}{dt} - g(x(t),y(t),t) \right) dt - \\ & - \mu_0
  (y(0) - y_0) 
\end{aligned}
$$

---

## Pontryagin's maximum principle

- Problem:
$$
\begin{aligned}
\max_{x,y} & \int_0^T F(x(t),y(t),t) dt \\
& \text{ s.t.} \\
    &  \frac{d y}{dt} = g(x(t),y(t),t) \forall t \in
        [0,T]  \\ 
        & y(0) = y_0
\end{aligned}
$$
- Hamiltonian
\[ H(x,y,\lambda,t) = F(x(t),y(t),t) + \lambda(t) g(x(t),y(t),t) \]
- If $x^*$ and $y^*$ are a local constrained maximizers, then there exists
  $\lambda^*(t)$ such that 
  $$
  \begin{aligned}
    [x]: && 0 = & \frac{\partial H}{\partial x}(x^*,y^*,\lambda^*,t)
    \\
    [y]: && -\frac{d\lambda}{dt}(t) = & \frac{\partial H}{\partial y}(x^*,y^*,\lambda^*,t) \\
    [\lambda]: && \frac{dy}{dt}(t) = & \frac{\partial H}{\partial
      \lambda}(x^*,y^*,\lambda^*,t)
  \end{aligned}
  $$

---

<!--- ##################################################### --->

# Examples

## Landlord housing investment

$$
  \begin{aligned}
    \max_{x,y} & \int_0^T p(t) y(t) - s(t) x(t) - c(t) x(t)^2 dt \\
    \text{s.t.} & \frac{dy}{dt} = x(t) \\
               & y(0) = y_0. 
  \end{aligned}
$$

- $y(t)$ units of housing
- $p(t)$ rental price
- $x(t)$ housing investment
- $s(t)$ price of housing, $c(t)x(t)^2$ adjustment cost

---
## Linear production and savings

$$
  \begin{aligned}
    \max_{s,k} & \int_0^T (1-s(t)) k(t) dt \\
    \text{ s.t. } & \frac{dk}{dt} = s(t)k(t) \\
    & k(0) = k_0 \\
    & k(t) \geq 0 \\
    & 0 \leq s(t) \leq 1
  \end{aligned}
$$

--- 

##  Inventory with costly storage

$$ 
  \begin{aligned}
    \max_{x,y} & p y_T - \int_0^T \left( c x(t)^2  + s y(t) \right)
    dt \\
    \text{ s.t. } 
    & \frac{dy}{dt} = x(t) \\
    & y(T) = y_T \\
    & y(0) = 0 \\
    & x(t) \geq 0 
  \end{aligned}
$$

<!-- ############################################################ --> 

#  Optimal growth

---

## Optimal growth

$$
\begin{aligned}
  \max_{c(t),k(t)} & \int_0^\infty e^{-\delta t} u(c(t)) dt \\
  \text{ s.t. } & \frac{dk}{dt} = f(k(t)) - \phi k(t) - c(t) \\
  & k(0) = k_0 \\
  & 0 \leq c(t) \leq f(k(t))
  \end{aligned}
$$

--- 

## Evolution of optimal c and k

- Constraint
$$ \frac{dk}{dt} = f(k(t)) - \phi k(t) - c(t) $$
- First order conditions $\Rightarrow$
$$  \frac{dc}{dt} = -\frac{u'(c)}{u''(c)} \left(f'(k(t)) - \phi -
\delta \right) $$
- Given $k(0)$ and a choice of $c(0)$, these equations determine
  $c(t)$ and $k(t)$
- Steady state: $dk/dt = dc/dt= 0$
$$ f'(k^{ss}) - \phi - \delta = 0 $$
$$ f(k^{ss}) - \phi k^{ss} - c^{ss} = 0 $$
- Stable path: given $k(0) = k_0$, there is unique $c(0)$ such that
  $c(t)$ and $k(t)$ reach the steady state

--- 

## Phase diagram
```{r phase, echo=FALSE, warning=FALSE}
source("../phase.R")

ord <- order(d$k)
fig <- ggplot(data=d) +
  #geom_segment(data=d,aes(x=k,y=c,yend=(c+dc),xend=(k+dk)),
  #             arrow=arrow(length=unit(0.1,"cm")),colour="gray") +
  geom_line(aes(x=k,y=zerodk),colour="red") +
  geom_vline(xintercept=kss,colour="blue") +
  annotate("text",label="dc/dt=0",colour="blue",x=kss,y=quantile(d$zerodk,0.25)) +
  annotate("text",label="dk/dt=0",colour="red",x=d$k[ord[100]],y=d$zerodk[ord[100]]) +
  xlim(min(k.grid),max(k.grid)) +
  ylim(min(c.grid),max(c.grid)) + theme_minimal() + ylab("c")
print(fig)
```

--- 

## Phase diagram
```{r phase2, echo=FALSE, warning=FALSE}
fig2 <- fig + 
  geom_segment(data=d,aes(x=k,y=c,yend=(c+dc),xend=(k+dk)),
               arrow=arrow(length=unit(0.1,"cm")),colour="gray") 
print(fig2)
```

--- 

## Phase diagram
```{r phase3, echo=FALSE, warning=FALSE}
i <- sum(stable$k<25)
fig3 <- fig2 + 
  geom_line(data=stable,aes(x=k,y=c),colour="black") +
  annotate("text",label="stable path",colour="black",x=stable$k[i],y=stable$c[i])
print(fig3)
```

--- 

# Contracting

---

## Contracting

$$
\begin{aligned}
  \max_{q(\theta),T(\theta)} & \int_{\theta_l}^{\theta_h} 
  \left[T(\theta) - cq(\theta)\right]
  f_\theta(\theta) d\theta  \\
  & \text{s.t.}  \\
  &\theta \nu\left(q(\theta)\right) - T(\theta) \geq 0  \forall
  \theta \\
  &\theta \nu\left(q(\theta)\right) - T(\theta) \geq
  \max_{\tilde{\theta}} \theta \nu\left(q(\tilde{\theta}) \right) -
  T(\tilde{\theta}) \forall \theta 
\end{aligned}
$$

---

## Contracting

$$
\begin{aligned}
  \max_{q(\theta),T(\theta)} & \int_{\theta_l}^{\theta_h} 
  \left[T(\theta) - cq(\theta)\right]
  f_\theta(\theta) d\theta  \\
  & \text{s.t.}  \\
  &\theta_l \nu\left(q(\theta_l)\right) - T(\theta_l) \geq 0 \\
  & \theta \nu'(q(\theta))q'(\theta) - T'(\theta) =  0 \\
  & \frac{dq(\theta)}{d\theta} \geq  0 
\end{aligned}
$$

---
